---
title: "Check correctness species list information against GBIF Backbone"
subtitle: "Workflow to check the correctness of the GBIF taxon keys of the LIFE RIPARIAS early alert tool"
author:
- Damiano Oldoni
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

# Setup 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)
```


Load libraries:

```{r message = FALSE}
library(readr) # to read txt files
library(here) # to work with paths
library(rgbif) # to interface to the Global Biodiversity Information Facility API
library(dplyr) # to do data wrangling
library(purrr) # to do functional programming
library(tidylog) # to get informative messages while doing data wrangling 
library(assertthat) # to run assertions (checks)

```

# Read source data

Create a data frame `input_data` from the export of the species list used in the LIFE RIPARIAS [early alert tool](https://alert.riparias.be/):

```{r read_species_list}
input_data <- read_csv(
  file = here("data", "reference", "Species-2023-07-10.csv")
) 
```

Preview data:

```{r}
input_data %>% head(n = 5)
```

# Harvest taxonomic information from GBIF

## Harvest taxonomic information from checklists

The LIFE RIPARIAS early alert tool includes species of the following checklists:

- [RIPARIAS target species list](https://www.gbif.org/dataset/fd004d9a-2ea4-4244-bb60-0df508d20a15) (`datasetKey`: `fd004d9a-2ea4-4244-bb60-0df508d20a15`)
- [List of Invasive Alien Species of Union concern](https://www.gbif.org/dataset/79d65658-526c-4c78-9d24-1870d67f8439) (`datasetKey`: `79d65658-526c-4c78-9d24-1870d67f8439`)

```{r get_taxa}
checklists_keys <- c("fd004d9a-2ea4-4244-bb60-0df508d20a15",
                     "79d65658-526c-4c78-9d24-1870d67f8439"
)
taxa_checklists <- purrr::map_dfr(checklists_keys, 
                                  function(x) {
                                    rgbif::name_usage(datasetKey = x)$data %>%
                                      filter(origin == "SOURCE")
                                  }
)
```

View taxa:

```{r view}
taxa_checklists
```

## Harvest taxonomic information from the GBIF Backbone Taxonomy

The  `nubKey` are the GBIF taxon keys of the taxa as defined in the [GBIF Backbone Taxonomy](https://www.gbif.org/dataset/d7dddbf4-2cf0-4f39-9b2a-bb099caae36c).

All taxa have a match to the backbone:

```{r check_match_backbone}
assertthat::assert_that(all(!is.na(taxa_checklists$nubKey)))
```

We harvest taxonomic information from the backbone, as these keys are the ones linked to the occurrences and so used in the early alert tool:

```{r harvest_from_backbone}
taxa <- purrr::map_df(
  taxa_checklists$nubKey, 
  function(x) {
    taxa_info <- rgbif::name_usage(x)$data
  }
) %>%
  distinct()
```

View taxonomic information from backbone:

```{r backbone_taxa}
taxa
```

## Patches

### Synonyms

If a synonym occurs, we will use the correspondent accepted taxon key from the backbone.

Synonyms:

```{r}
synonyms <- taxa %>%
  filter(!taxonomicStatus %in% c("ACCEPTED", "DOUBTFUL"))
synonyms
```

Get taxonomic information of the correspondent accepted taxa:

```{r accepted}
accepted <- NULL
if (nrow(synonyms) > 0) {
  accepted <- purrr::map_df(
    synonyms$acceptedKey, 
    function(x) {
      taxa_info <- rgbif::name_usage(x)$data
    }
  )
}
```

View information about the accepted taxa (´NULL´ if no synonyms are detected):

```{r}
accepted
```

## Higher rank

The Asian hornet (_Vespa velutina_) is mentioned in the List of Invasive Alien Species of Union concern as a subspecies: _Vespa velutina nigrithorax_:

```{r asian_hornet_subspecies}
asian_hornet_subspecies <- taxa %>%
  filter(rank == "SUBSPECIES")
asian_hornet_subspecies
```

However, the vast majority of observations are at species level, i.e. _Vespa velutina_:

```{r get_info_at_species_level}
asian_hornet_species_key <- asian_hornet_subspecies$speciesKey
asian_hornet_species <- rgbif::name_usage(asian_hornet_species_key)$data
asian_hornet_species
```

# Combine information

The early alert tool uses latin names as defined in the checklists (column `canonicalName`) and GBIF taxon keys of the accepted taxa as defined in the GBIF Backbone:

```{r combine_names_gbif_keys}
alert_species <- taxa_checklists %>%
  distinct(canonicalName, nubKey) %>%
  # add GBIF Backbone information
  left_join(taxa %>% select(key, acceptedKey),
            by = c("nubKey" = "key")) %>%
  # remove synonyms whose accepted taxa are in the list
  filter(is.na(acceptedKey) | !acceptedKey %in% taxa_checklists$nubKey)
if (!is.null(accepted)) {
  alert_species <- alert_species %>%
    left_join(accepted %>%
                select(key),
              by = c("acceptedKey" = "key")) %>%
    mutate(key = if_else(!is.na(acceptedKey),
                         acceptedKey,
                         nubKey))
}
alert_species <- alert_species %>% select(-c(nubKey, acceptedKey))
```

Use species information for the Asian hornet:

```{r}
alert_species <- alert_species %>%
  # remove subspecies Vespa velutina nigrithorax
  filter(key != asian_hornet_subspecies$nubKey) %>%
  # add species Vespa velutina
  union(asian_hornet_species %>% select(canonicalName, key))
```

Show alert species:

```{r alert_species}
alert_species
```

## Save output

Save the list of names and GBIF taxon keys:

```{r write_alert_species}
write_csv(alert_species, 
          file = here::here("data", "processed", "alert_species.csv"),
          na = ""
)
```

# Check correctness species list of the LIFE RIPARIAS early alert tool

We can now check the correctness of the information as defined in the early alert tool (species list in `./data/reference`)


Check species names:

```{r check_names_latin}
assertthat::assert_that(all(alert_species$canonicalName %in% input_data$name))
```

Check GBIF taxon keys:

```{r check_keys}
assertthat::assert_that(all(input_data$gbif_taxon_key %in% alert_species$key))
```
